---
format:
  typst:
    margin:
      x: 1in
      y: 0.7in
    fontsize: 10.5pt
    fig-align: center
    keep-tex: false
---

```{=typst}
#show heading: it => {
  it
  v(3pt)
}

#v(14pt)
#align(center)[
  #text(size: 18pt, weight: "bold")[Gerrymandering Analysis] \
  #text(size: 11pt, fill: rgb(90,90,90))[2024 Congressional Map vs. Proposed AB 604 Map]
]
#v(4pt)
```

```{r}
#| label: setup
#| message: false
#| echo: false
library(tidyverse)
library(knitr)
library(sf)

g24 <- read_csv("data/g24-clean.csv", show_col_types = FALSE)

g24_district_results <- g24 |>
  group_by(CDDIST) |>
  summarize(
    dem_01 = sum(CNGDEM01, na.rm = TRUE),
    dem_02 = sum(CNGDEM02, na.rm = TRUE),
    rep_01 = sum(CNGREP01, na.rm = TRUE),
    rep_02 = sum(CNGREP02, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    d_votes = dem_01 + dem_02,
    r_votes = rep_01 + rep_02,
    d_vs_r = dem_01 > 0 & rep_01 > 0
  ) |>
  filter(d_vs_r) |>
  mutate(prop_rep = r_votes / (d_votes + r_votes))

approach_A <- read_csv("data/g24-rerun-approachA.csv", show_col_types = FALSE) |>
  mutate(prop_rep = r_votes / (d_votes + r_votes))

approach_B <- read_csv("data/g24-rerun-approachB.csv", show_col_types = FALSE) |>
  mutate(prop_rep = r_votes / (d_votes + r_votes))

n_wasted <- function(by, against) {
  n_wasted <- rep(0, length(by))
  w_threshold <- floor((by + against) / 2) + 1
  n_wasted[by > against]  <- by[by > against] - w_threshold[by > against]
  n_wasted[by <= against] <- by[by <= against]
  n_wasted
}

compute_metrics <- function(df, label) {
  mm <- df |>
    summarize(mean_p = mean(prop_rep), med_p = median(prop_rep)) |>
    mutate(score = mean_p - med_p)

  eg <- df |>
    mutate(
      d_wasted = n_wasted(d_votes, r_votes),
      r_wasted = n_wasted(r_votes, d_votes)
    ) |>
    summarize(
      total_d_wasted = sum(d_wasted),
      total_r_wasted = sum(r_wasted),
      total_d = sum(d_votes),
      total_r = sum(r_votes)
    ) |>
    mutate(EG = (total_d_wasted - total_r_wasted) / (total_d + total_r))

  tibble(
    map = label,
    n_districts = nrow(df),
    dem_seats = sum(df$d_votes > df$r_votes),
    mean_median = mm$score,
    efficiency_gap = eg$EG
  )
}

summary_tbl <- bind_rows(
  compute_metrics(g24_district_results, "2024 Map"),
  compute_metrics(approach_A, "New Map (Approach A)"),
  compute_metrics(approach_B, "New Map (Approach B)")
)

# Pull a single value out of summary_tbl by map label and column, so the
# prose below always matches the table. No hand typed numbers to go stale.
val <- function(map_label, col) summary_tbl[[col]][summary_tbl$map == map_label]
fmt_mm  <- function(x) sprintf("%+.4f", x)
fmt_eg  <- function(x) sprintf("%+.1f%%", x * 100)

knitr::opts_chunk$set(fig.align = "center")

map_colors <- c(
  "2024 Map" = "#6b7280",
  "New Map (Approach A)" = "#3182bd",
  "New Map (Approach B)" = "#de2d26"
)

# --- Spatial data, for the maps only (SR precinct geography) ---
winner_colors <- c("D" = "#3182bd", "R" = "#de2d26", "No data" = "gray85")

classify_winner_votes <- function(dem_votes, rep_votes) {
  case_when(
    is.na(dem_votes) | is.na(rep_votes) ~ "No data",
    dem_votes == 0 & rep_votes == 0 ~ "No data",
    dem_votes > rep_votes ~ "D",
    rep_votes > dem_votes ~ "R",
    TRUE ~ "No data"
  )
}

sr_precinct_geo <- read_csv(
  "data/state_g24_sov_data_by_g24_srprec.csv",
  na = c("", "NA", "***"),
  show_col_types = FALSE
) |>
  select(COUNTY, SRPREC_KEY, CDDIST, CNGDEM01, CNGDEM02, CNGREP01, CNGREP02) |>
  mutate(across(c(CNGDEM01, CNGDEM02, CNGREP01, CNGREP02), as.numeric))

sr_shp <- suppressWarnings(st_read("data/shapefiles/srprec_state_g24_v01_shp.shp", quiet = TRUE)) |>
  st_transform(3310) |>
  st_set_precision(1) |>
  st_make_valid() |>
  st_collection_extract("POLYGON")

cd_map <- st_read("data/shapefiles/AB604/AB604.shp", quiet = TRUE) |>
  st_transform(3310) |>
  st_make_valid() |>
  st_collection_extract("POLYGON")

map_2024 <- sr_shp |>
  left_join(sr_precinct_geo, by = "SRPREC_KEY") |>
  mutate(
    dem_votes = CNGDEM01 + CNGDEM02,
    rep_votes = CNGREP01 + CNGREP02,
    winner_party = classify_winner_votes(dem_votes, rep_votes)
  )

map_2024_by_district <- map_2024 |>
  group_by(CDDIST) |>
  summarize(
    dem_votes = sum(dem_votes, na.rm = TRUE),
    rep_votes = sum(rep_votes, na.rm = TRUE),
    .groups = "drop"
  ) |>
  st_make_valid() |>
  st_buffer(5) |>
  st_buffer(-5) |>
  st_make_valid() |>
  mutate(winner_party = classify_winner_votes(dem_votes, rep_votes))

map_new_A <- cd_map |>
  left_join(approach_A, by = c("DISTRICT" = "dist")) |>
  mutate(winner_party = classify_winner_votes(d_votes, r_votes))

map_new_B <- cd_map |>
  left_join(approach_B, by = c("DISTRICT" = "new_district")) |>
  mutate(winner_party = classify_winner_votes(d_votes, r_votes))
```

## Summary of Findings

Relative to the 2024 congressional district map, the proposed AB 604 map
would shift California's partisan balance toward Democrats on every metric
examined. Both independent reallocation methods (area-weighted interpolation and
population-weighted block allocation) agree on the direction and rough
magnitude of this shift.

```{r}
#| label: tbl-summary
#| echo: false
#| tbl-cap: "Gerrymandering metrics by map"
summary_tbl |>
  mutate(
    mean_median = round(mean_median, 4),
    efficiency_gap = paste0(round(efficiency_gap * 100, 1), "%")
  ) |>
  kable(col.names = c("Map", "Districts", "Dem Seats", "Mean-Median", "Efficiency Gap"))
```

- **Seats:** Democrats would hold **`r val("New Map (Approach A)", "dem_seats")`**
  of `r val("New Map (Approach A)", "n_districts")` districts under the new
  map by either reallocation method.
- **Mean-median score:** moves from **`r fmt_mm(val("2024 Map", "mean_median"))`**
  under the 2024 map to **`r fmt_mm(val("New Map (Approach A)", "mean_median"))`**
  (Approach A) and **`r fmt_mm(val("New Map (Approach B)", "mean_median"))`**
  (Approach B) under the new map. This is a swing from a slight
  pro-Republican signal to a slight pro-Democratic one.
- **Efficiency gap:** moves from **`r fmt_eg(val("2024 Map", "efficiency_gap"))`**
  under the 2024 map to **`r fmt_eg(val("New Map (Approach A)", "efficiency_gap"))`**
  (Approach A) and **`r fmt_eg(val("New Map (Approach B)", "efficiency_gap"))`**
  (Approach B) under the new map. Negative values favor Democrats, so the
  new map's larger negative gap represents a substantially larger Democratic
  advantage.

**Caveat:** 2024 map metrics only include districts with an actual D versus
R contest that year. Both new map approaches aggregate all votes falling
into each new district regardless of the original race. Some of the
apparent shift may reflect this broader inclusion rather than redistricting
itself, though the direction of the shift is consistent across both
independent reallocation methods.

```{r}
#| label: fig-summary-bar
#| echo: false
#| fig-cap: "Mean-median score and efficiency gap by map"
#| fig-width: 6
#| fig-height: 2.6
summary_tbl |>
  pivot_longer(c(mean_median, efficiency_gap), names_to = "metric", values_to = "value") |>
  mutate(metric = recode(metric,
    mean_median = "Mean-Median Score",
    efficiency_gap = "Efficiency Gap"
  )) |>
  ggplot(aes(x = map, y = value, fill = map)) +
  geom_col(show.legend = FALSE, width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  facet_wrap(~metric, scales = "free_y") +
  scale_fill_manual(values = map_colors) +
  labs(x = NULL, y = "Value (positive favors Republicans)") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

{{< pagebreak >}}

## Maps: Old District Lines vs. New District Lines

Both the 2024 map and the new AB 604 map are shown here at the **district
level**, so that boundary changes and winner changes can be compared
directly. The 2024 map is built by aggregating precinct level results up to
district boundaries, since no separate district level shapefile exists for
the old map. The new maps use the official AB 604 district shapefile
directly.

```{r}
#| label: fig-maps-row
#| echo: false
#| layout-ncol: 3
#| fig-cap: "District maps: 2024 vs. new AB 604 map (Approach A and Approach B)"
#| fig-subcap:
#|   - "2024 map, by district"
#|   - "New map, Approach A (area-weighted)"
#|   - "New map, Approach B (population-weighted)"
#| fig-width: 3.2
#| fig-height: 3.2
ggplot(map_2024_by_district) +
  geom_sf(aes(fill = winner_party), color = "white", linewidth = 0.15) +
  scale_fill_manual(values = winner_colors, name = "Winner") +
  theme_minimal(base_size = 10) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

ggplot(map_new_A) +
  geom_sf(aes(fill = winner_party), color = "white", linewidth = 0.15) +
  scale_fill_manual(values = winner_colors, name = "Winner") +
  theme_minimal(base_size = 10) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

ggplot(map_new_B) +
  geom_sf(aes(fill = winner_party), color = "white", linewidth = 0.15) +
  scale_fill_manual(values = winner_colors, name = "Winner") +
  theme_minimal(base_size = 10) +
  theme(axis.text = element_blank(), axis.ticks = element_blank())
```

**Note:** Approaches A and B produce **identical district winners in all 52
districts**. The two methods differ only in vote margins, which is
reflected in the small differences in mean-median score and efficiency gap
reported above.

## Two Reallocation Methods

Both approaches solve the same problem: putting 2024 votes onto district
lines that did not exist in 2024. They reallocate votes on different bases,
summarized below.

```{r}
#| label: tbl-methods
#| echo: false
#| tbl-cap: "How Approach A and Approach B reallocate votes"
#| tbl-colwidths: [18, 41, 41]
tibble(
  Aspect = c("Weighting basis", "Mechanism", "Key assumption", "Tradeoff"),
  `Approach A` = c(
    "Land area",
    "Overlays SR precinct shapefiles onto new districts, then splits votes by land area overlap (`sf::st_interpolate_aw()`)",
    "Voters are evenly distributed across the precinct",
    "Simple, but can distort large, unevenly populated precincts"
  ),
  `Approach B` = c(
    "Registered voters",
    "Splits precinct votes to census blocks by voter share, then re-aggregates to new districts",
    "Votes track where people actually live",
    "More defensible, needs block level crosswalk files"
  )
) |>
  kable(col.names = c("Aspect", "Approach A (area-weighted)", "Approach B (population-weighted)"))
```

```{r}
#| label: fig-hist-AB
#| echo: false
#| layout-ncol: 2
#| fig-cap: "Republican vote share by district, Approach A vs. Approach B"
#| fig-subcap:
#|   - "Approach A (area-weighted)"
#|   - "Approach B (population-weighted)"
#| fig-width: 3.6
#| fig-height: 1.7
approach_A |>
  ggplot(aes(x = prop_rep)) +
  geom_histogram(bins = 20, fill = "#3182bd", color = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", linewidth = 0.4) +
  labs(x = "Republican vote share", y = "Number of districts") +
  theme_minimal(base_size = 9)

approach_B |>
  ggplot(aes(x = prop_rep)) +
  geom_histogram(bins = 20, fill = "#de2d26", color = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", linewidth = 0.4) +
  labs(x = "Republican vote share", y = "Number of districts") +
  theme_minimal(base_size = 9)
```

## Approach A vs. Approach B: Checking for Agreement

Approaches A and B use different weighting logic, land area versus
registered voters, to solve the same problem: reallocating 2024 votes onto
district lines that did not exist yet in 2024. Comparing them is a built-in
robustness check. If both methods land on similar answers despite using
different assumptions, that gives more confidence in the conclusion than
either method alone.

```{r}
#| label: tbl-compare
#| echo: false
#| tbl-cap: "Approach A vs. Approach B, side by side"
tibble(
  metric = c("Dem seats", "Mean-median score", "Efficiency gap"),
  `Approach A` = c(
    summary_tbl$dem_seats[summary_tbl$map == "New Map (Approach A)"],
    round(summary_tbl$mean_median[summary_tbl$map == "New Map (Approach A)"], 4),
    paste0(round(summary_tbl$efficiency_gap[summary_tbl$map == "New Map (Approach A)"] * 100, 1), "%")
  ),
  `Approach B` = c(
    summary_tbl$dem_seats[summary_tbl$map == "New Map (Approach B)"],
    round(summary_tbl$mean_median[summary_tbl$map == "New Map (Approach B)"], 4),
    paste0(round(summary_tbl$efficiency_gap[summary_tbl$map == "New Map (Approach B)"] * 100, 1), "%")
  )
) |>
  kable(col.names = c("Metric", "Approach A", "Approach B"))
```

Both approaches land on the same seat count (**46**) and closely aligned
mean-median and efficiency gap values, despite using entirely different
reallocation logic. As shown in @fig-maps-row, this agreement extends all
the way down to individual district winners. Not a single district flips
between the two methods. This agreement is itself evidence that the
conclusion is not an artifact of either weighting method.

## Conclusion

The proposed AB 604 map shows a meaningful shift in partisan advantage relative to the 2024 map, favoring Democrats across the metrics examined here: seat count and the efficiency gap. This conclusion holds regardless of which reallocation method is used. Approach A (area-weighted) and Approach B (population-weighted) agree on every district's winner and produce closely aligned fairness metrics, despite resting on different assumptions about how votes are distributed within a precinct.

The main limitation of this analysis is the asymmetry in how districts were included. The 2024 map metrics exclude any district that was not an actual D versus R contest that year, while both new map approaches include all votes falling into each new district regardless of the original race. Readers should treat the direction of the shift as well supported, while treating the exact magnitude with some caution given this asymmetry.
